---
title: "Final RNA-seq Project"
author: "Grace Scanlan, Tavi Mangoma, Orlando Pender"
date: "2026"
output: html_document
---
# RNASeq R pipeline
# Normalize Counts
### Install packages if not already installed and import libraries
```{r}
if (!require("BiocManager")) install.packages("BiocManager")
if (!require("DESeq2")) BiocManager::install("DESeq2")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("apeglm")) BiocManager::install("apeglm")
if (!require("org.Hs.eg.db")) BiocManager::install("org.Hs.eg.db")
if (!require("AnnotationDbi")) BiocManager::install("AnnotationDbi")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("ggrepel", quietly = TRUE)) install.packages("ggrepel")
if (!requireNamespace("pheatmap", quietly = TRUE)) install.packages("pheatmap")
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
if (!requireNamespace("pheatmap", quietly = TRUE)) install.packages("pheatmap")
if (!requireNamespace("matrixStats", quietly = TRUE)) install.packages("matrixStats")
if (!requireNamespace("UpSetR", quietly = TRUE)) install.packages("UpSetR")
if (!requireNamespace("patchwork", quietly = TRUE)) install.packages("patchwork")
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")

library(DESeq2)
library(tidyverse)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(here)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(here)
library(pheatmap)
library(matrixStats)
library(UpSetR)
library(patchwork) 
library(tidyr)
library(dplyr)
```
### set paths to project root and counts directory
```{r}
counts_path <- here("counts", "gene_counts.txt")
meta_path <- here("config", "metadata.tsv")
normalized_dir <- here("normalized_counts")
```
### create normalized counts directory if it doesn't exist
```{r}
if (!dir.exists(normalized_dir)) {
  dir.create(normalized_dir, recursive = TRUE)
}
```
### Load gene counts
```{r}
counts <- read.table(counts_path,
                     header = TRUE,
                     row.names = 1,
                     comment.char = "#")
```
### Keep only sample columns (drop featureCounts annotation columns)
```{r}
counts <- counts[, 6:ncol(counts)]
```
### Clean counts column names:  HISAT2.SRRxxxx.sorted(.bam) -> SRRxxxx
```{r}
colnames(counts) <- gsub("^HISAT2\\.", "", colnames(counts))
colnames(counts) <- gsub("\\.sorted(\\.bam)?$", "", colnames(counts))
colnames(counts) <- trimws(colnames(counts))
```
### Look at the first few rows
```{r}
head(counts)
```
### Load metadata
Metadata was created manually.
```{r}
metadata <- read.table(meta_path, header = TRUE, sep = "\t")
```
### Set metadata row names to sample IDs (and trim whitespace)
```{r}
metadata$sample_id <- trimws(metadata$sample_id)
rownames(metadata) <- metadata$sample_id
```
### Sanity Checks
```{r}
cat("Count matrix dimensions:\n")
print(dim(counts))
cat("\nMetadata dimensions:\n")
print(dim(metadata))

cat("\nPreview of count matrix:\n")
print(head(counts[, 1:min(5, ncol(counts))]))

cat("\nPreview of metadata:\n")
print(head(metadata))

missing <- setdiff(rownames(metadata), colnames(counts))
cat("\nSamples in metadata but not counts:\n")
print(missing)

stopifnot(length(missing) == 0)
```
### reorder/subset
```{r}
counts <- counts[, rownames(metadata)]
```
### Create DESeq2 dataset
```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = metadata,
  design = ~ tumor_type
)
```
### Make sure condition is a factor
```{r}
dds$tumor_type <- factor(dds$tumor_type)
cat("\nTumor type levels:\n")
print(levels(dds$tumor_type))
```
### remove very low count genes
```{r}
dds <- dds[rowSums(counts(dds)) > 10, ]

cat("\nGenes remaining after filtering:\n")
print(nrow(dds))
```
### Normalize counts
```{r}
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized = TRUE)
```
### compare raw vs normalized counts for a few genes
```{r}
head(counts(dds)[1:5, 1:5])
head(normalized_counts[1:5, 1:5])
```
### Save normalized counts to file
```{r}
out_file <- file.path("..", "normalized_counts", "normalized_gene_counts.csv")
write.csv(normalized_counts, file = out_file, row.names = TRUE)
```
## Before vs After Normalization plots
### Boxplot of raw counts
```{r}
boxplot(log2(counts + 1),
        main = "Raw Counts",
        xlab = "Samples",
        ylab = "Log2(Counts + 1)",
        las = 2)
```
### Boxplot of normalized counts
```{r}
boxplot(log2(normalized_counts + 1),
        main = "Normalized Counts",
        xlab = "Samples",
        ylab = "Log2(Normalized Counts + 1)",
        las = 2)
```
### Save boxplots to file
```{r}
boxplot_file <- file.path("..", "normalized_counts", "boxplots.pdf")
pdf(boxplot_file)

boxplot(log2(counts + 1),
        main = "Raw Counts",
        las = 2)

boxplot(log2(normalized_counts + 1),
        main = "Normalized Counts",
        las = 2)

dev.off()
```
### Overlay Density Plot raw vs normalized
```{r}
pdf(here("normalized_counts", "raw_vs_normalized_density.pdf"))

raw_vec  <- log2(as.vector(as.matrix(counts)) + 1)
norm_vec <- log2(as.vector(as.matrix(normalized_counts)) + 1)
```
### Drop any weird values just in case 
```{r}
raw_vec  <- raw_vec[is.finite(raw_vec)]
norm_vec <- norm_vec[is.finite(norm_vec)]

plot(density(raw_vec),
     col = "red",
     main = "Raw vs Normalized Counts",
     xlab = "Log2(counts + 1)")

lines(density(norm_vec), col = "blue")

legend("topright",
       legend = c("Raw", "Normalized"),
       col = c("red", "blue"),
       lwd = 2)

dev.off()
```
# PCA of normalized counts
### PCA outputs
```{r}
pca_dir <- here("output", "pca")

if (!dir.exists(pca_dir)) {
  dir.create(pca_dir, recursive = TRUE)
}
```
### Variance stabilizing transformation (VST) for PCA
```{r}
vsd <- vst(dds, blind = FALSE)
```
### PCA plot
```{r}
pca_plot <- plotPCA(vsd, intgroup = "tumor_type") +
  ggtitle("PCA Plot") +
  theme_minimal()

ggsave(
  filename = file.path(pca_dir, "pca_plot.pdf"),
  plot = pca_plot,
  width = 8,
  height = 6
)
```
### Subset PCA plot (remove one normal sample to see if it improves separation)
```{r}
dds_subset <- dds[, colnames(dds) != "SRR36750771"]

vsd_subset <- vst(dds_subset, blind = FALSE)

pca_subset_plot <- plotPCA(vsd_subset, intgroup = "tumor_type") +
  ggtitle("PCA Plot (Subset: One Normal Sample Removed)") +
  theme_minimal()

ggsave(
  filename = file.path(pca_dir, "pca_subset.pdf"),
  plot = pca_subset_plot,
  width = 8,
  height = 6
)
```
# Heatmaps
### Create output directory for heatmaps
```{r}
heatmap_dir <- here("output", "heatmaps")

if (!dir.exists(heatmap_dir)) {
  dir.create(heatmap_dir, recursive = TRUE)
}
```
## Sample distance heatmap
```{r}
mat <- assay(vsd)

sample_dists <- dist(t(mat))
sample_dist_matrix <- as.matrix(sample_dists)

rownames(sample_dist_matrix) <- colnames(mat)
colnames(sample_dist_matrix) <- colnames(mat)

pdf(file.path(heatmap_dir, "sample_distance_heatmap.pdf"), width = 8, height = 8)

pheatmap(
  sample_dist_matrix,
  clustering_distance_rows = sample_dists,
  clustering_distance_cols = sample_dists,
  main = "Sample Distance Heatmap"
)

dev.off()
```
## Top significant genes heatmap
### keep only significant genes for heatmap
```{r}
sig_only <- res_df[!is.na(res_df$padj) & res_df$padj < 0.05, ]
```
### sort by adjusted p-value
```{r}
sig_only <- sig_only[order(sig_only$padj), ]
```
### choose top genes
```{r}
top_n <- 20
top_genes <- sig_only$gene_id[1:min(top_n, nrow(sig_only))]
```
### extract vst matrix values for those genes
```{r}
mat_top <- assay(vsd)[top_genes, , drop = FALSE]
```
### scale by gene (row)
```{r}
mat_scaled <- t(scale(t(mat_top)))
```
### remove rows with non-finite values after scaling
```{r}
keep_rows <- rowSums(is.na(mat_scaled) | is.infinite(mat_scaled)) == 0
mat_scaled <- mat_scaled[keep_rows, , drop = FALSE]
```
### build row labels using gene symbols when available
```{r}
gene_lookup <- res_df
rownames(gene_lookup) <- gene_lookup$gene_id

row_labels <- gene_lookup[rownames(mat_scaled), "gene_symbol"]
row_labels[is.na(row_labels) | row_labels == ""] <- rownames(mat_scaled)

rownames(mat_scaled) <- row_labels
```
### plot heatmap
```{r}
annotation_col <- as.data.frame(colData(vsd))[, "tumor_type", drop = FALSE]

pdf(file.path(heatmap_dir, "top_significant_genes_heatmap.pdf"), width = 8, height = 10)

pheatmap(
  mat_scaled,
  annotation_col = annotation_col,
  show_rownames = TRUE,
  main = "Top Significant Genes Heatmap",
  scale = "none"
)

dev.off()
```
# Differential Expression Analysis
### DEG outputs
```{r}
deg_dir <- here("output", "advanced_vs_normal")

if (!dir.exists(deg_dir)) {
  dir.create(deg_dir, recursive = TRUE)
}
```
### Differential expression analysis using DESeq2
```{r}
dds <- DESeq(dds)
```
## Advanced Tumor vs Normal Tissue
### Make sure the contrast matches reference level setup
```{r}
res <- results(dds, contrast = c("tumor_type", "Advanced_Tumor", "Normal"))
```
### Clean result Table
```{r}
res_df <- as.data.frame(res)
res_df$gene_id <- rownames(res_df)

res_df <- res_df[!is.na(res_df$padj) & !is.na(res_df$log2FoldChange), ]
```
### Add gene symbols and names 
```{r}
res_df$gene_symbol <- mapIds(
  org.Hs.eg.db,
  keys = res_df$gene_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

res_df$gene_name <- mapIds(
  org.Hs.eg.db,
  keys = res_df$gene_id,
  column = "GENENAME",
  keytype = "ENSEMBL",
  multiVals = "first"
)
```
### DE Summary Counts
```{r}
advanced_genes_tested <- nrow(res_df)
advanced_sig_count <- sum(res_df$padj < 0.05, na.rm = TRUE)
advanced_up_count <- sum(res_df$padj < 0.05 & res_df$log2FoldChange > 1, na.rm = TRUE)
advanced_down_count <- sum(res_df$padj < 0.05 & res_df$log2FoldChange < -1, na.rm = TRUE)

cat("Genes tested:", advanced_genes_tested, "\n")
cat("Significant genes (FDR < 0.05):", advanced_sig_count, "\n")
cat("Upregulated genes:", advanced_up_count, "\n")
cat("Downregulated genes:", advanced_down_count, "\n")
```
### Regulation groups
```{r}
res_df$Regulation <- "NS"
res_df$Regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$Regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"
```
### Top 10 genes for labeling
```{r}
top_genes <- res_df[order(res_df$padj), ][1:10, ]
top_genes$label <- ifelse(
  is.na(top_genes$gene_symbol) | top_genes$gene_symbol == "",
  top_genes$gene_id,
  top_genes$gene_symbol
)
```
### Export DEG results
```{r}
deg_results <- res_df[, c("gene_id", "log2FoldChange", "pvalue", "padj", "Regulation", "gene_symbol", "gene_name")]
colnames(deg_results) <- c(
  "Gene_ID", "log2FC", "p_value", "adjusted_p_value",
  "Regulation", "Gene_Symbol", "Gene_Name"
)

write.csv(
  deg_results,
  file.path(deg_dir, "DEG_results_advanced_vs_normal.csv"),
  row.names = FALSE
)
```
### Number of significant genes (FDR < 0.05) and directionality 
```{r}
advanced_num_sig <- sum(res_df$padj < 0.05, na.rm = TRUE)
advanced_up_genes <- sum(res_df$padj < 0.05 & res_df$log2FoldChange > 0, na.rm = TRUE)
advanced_down_genes <- sum(res_df$padj < 0.05 & res_df$log2FoldChange < 0, na.rm = TRUE)
```
### Volcano plot
```{r}
volcano_plot_advanced <- ggplot(
  res_df,
  aes(x = log2FoldChange, y = -log10(padj), color = Regulation)
) +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    data = top_genes,
    aes(label = label),
    size = 3,
    max.overlaps = 100
  ) +
  scale_color_manual(values = c("Down" = "blue", "NS" = "gray", "Up" = "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  annotate(
  "text",
  x = 1.5,
  y = 26,
  label = paste0(
    "Significant DEGs: ", advanced_num_sig, "\n",
    "Upregulated: ", advanced_up_genes, "\n",
    "Downregulated: ", advanced_down_genes
  ),
  hjust = 0,
  vjust = 1,
  size = 3.6
) +
coord_cartesian(clip = "off") +
theme(
  plot.margin = margin(10, 40, 10, 10)
) +
  theme_minimal() +
  labs(
    title = "Volcano Plot",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value"
  )
ggsave(
  filename = file.path(deg_dir, "volcano_plot_advanced_vs_normal.pdf"),
  plot = volcano_plot_advanced,
  width = 8,
  height = 6
)
```
### MA plot
```{r}
ma_plot_advanced <- ggplot(
  res_df,
  aes(x = baseMean, y = log2FoldChange, color = Regulation)
) +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    data = top_genes,
    aes(label = label),
    size = 3,
    max.overlaps = 100
  ) +
  scale_x_log10() +
  scale_color_manual(values = c("Down" = "blue", "NS" = "gray", "Up" = "red")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "MA Plot",
    x = "Mean Expression (log scale)",
    y = "Log2 Fold Change"
  )

ggsave(
  filename = file.path(deg_dir, "ma_plot_advanced_vs_normal.pdf"),
  plot = ma_plot_advanced,
  width = 8,
  height = 6
)
```
### Top 10 DEG table for report 
```{r}
top10_deg <- head(deg_results[order(deg_results$adjusted_p_value), ], 10)
print(top10_deg)

write.csv(
  top10_deg,
  file.path(deg_dir, "top10_DEGs_advanced_vs_normal.csv"),
  row.names = FALSE
)
```
# Villous Adenoma vs Normal Tissue
### Set output directory for this analysis
```{r}
output_dir <- here("output", "villous_vs_normal")

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```
### Differential expression analysis using DESeq2
```{r}
dds <- DESeq(dds)
res <- results(dds, contrast = c("tumor_type", "Villous_Adenoma", "Normal"))
```
### Clean result Table
```{r}
res_df <- as.data.frame(res)
res_df$gene_id <- rownames(res_df)

res_df <- res_df[!is.na(res_df$padj) & !is.na(res_df$log2FoldChange), ]
```
### Add gene symbols and names
```{r}
res_df$gene_symbol <- mapIds(
  org.Hs.eg.db,
  keys = res_df$gene_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

res_df$gene_name <- mapIds(
  org.Hs.eg.db,
  keys = res_df$gene_id,
  column = "GENENAME",
  keytype = "ENSEMBL",
  multiVals = "first"
)
```
### DE Summary Counts
```{r}
villous_genes_tested <- nrow(res_df)
villous_sig_count <- sum(res_df$padj < 0.05, na.rm = TRUE)
villous_up_count <- sum(res_df$padj < 0.05 & res_df$log2FoldChange > 1, na.rm = TRUE)
villous_down_count <- sum(res_df$padj < 0.05 & res_df$log2FoldChange < -1, na.rm = TRUE)

cat("Genes tested:", villous_genes_tested, "\n")
cat("Significant genes (FDR < 0.05):", villous_sig_count, "\n")
cat("Upregulated genes:", villous_up_count, "\n")
cat("Downregulated genes:", villous_down_count, "\n")
```
### Regulation groups
```{r}
res_df$Regulation <- "NS"
res_df$Regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$Regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"
```
### Top 10 genes for labeling
```{r}
top_genes <- res_df[order(res_df$padj), ][1:10, ]

top_genes$label <- ifelse(
  is.na(top_genes$gene_symbol) | top_genes$gene_symbol == "",
  top_genes$gene_id,
  top_genes$gene_symbol
)
```
### Export DEG results
```{r}
deg_results <- res_df[, c("gene_id", "log2FoldChange", "pvalue", "padj", "Regulation", "gene_symbol", "gene_name")]
colnames(deg_results) <- c(
  "Gene_ID", "log2FC", "p_value", "adjusted_p_value",
  "Regulation", "Gene_Symbol", "Gene_Name"
)

write.csv(
  deg_results,
  file.path(output_dir, "DEG_results_villous_vs_normal.csv"),
  row.names = FALSE
)
```
### Count significant DEGs
```{r}
villous_num_sig <- sum(res_df$padj < 0.05, na.rm = TRUE)
villous_up_genes <- sum(res_df$padj < 0.05 & res_df$log2FoldChange > 0, na.rm = TRUE)
villous_down_genes <- sum(res_df$padj < 0.05 & res_df$log2FoldChange < 0, na.rm = TRUE)
```
### Volcano plot
```{r}
volcano_plot_villous <- ggplot(
  res_df,
  aes(x = log2FoldChange, y = -log10(padj), color = Regulation)
) +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    data = top_genes,
    aes(label = label),
    size = 3,
    max.overlaps = 100
  ) +
  scale_color_manual(values = c("Down" = "blue", "NS" = "gray", "Up" = "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  annotate(
  "text",
  x = 1.5,
  y = 26,
  label = paste0(
    "Significant DEGs: ", villous_num_sig, "\n",
    "Upregulated: ", villous_up_genes, "\n",
    "Downregulated: ", villous_down_genes
  ),
  hjust = 0,
  vjust = 1,
  size = 3.6
) +
coord_cartesian(clip = "off") +
theme(
  plot.margin = margin(10, 40, 10, 10)
) +
  theme_minimal() +
  labs(
    title = "Volcano Plot: Villous Adenoma vs Normal",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value"
  )

ggsave(
  filename = file.path(output_dir, "volcano_plot_villous_vs_normal.pdf"),
  plot = volcano_plot_villous,
  width = 8,
  height = 6
)
```
### MA plot
```{r}
ma_plot_villous <- ggplot(
  res_df,
  aes(x = baseMean, y = log2FoldChange, color = Regulation)
) +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    data = top_genes,
    aes(label = label),
    size = 3,
    max.overlaps = 100
  ) +
  scale_x_log10() +
  scale_color_manual(values = c("Down" = "blue", "NS" = "gray", "Up" = "red")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "MA Plot: Villous Adenoma vs Normal",
    x = "Mean Expression (log scale)",
    y = "Log2 Fold Change"
  )

ggsave(
  filename = file.path(output_dir, "ma_plot_villous_vs_normal.pdf"),
  plot = ma_plot_villous,
  width = 8,
  height = 6
)
```
### Top 10 DEG table
```{r}
top10_deg <- head(deg_results[order(deg_results$adjusted_p_value), ], 10)
print(top10_deg)

write.csv(
  top10_deg,
  file.path(output_dir, "top10_DEGs_villous_vs_normal.csv"),
  row.names = FALSE
)
```
# Tubular Adenoma vs Normal Tissue
### Set output directory for this analysis
```{r}
output_dir <- here("output", "tubular_vs_early_stage")
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```
### Differential expression analysis using DESeq2
```{r}
dds <- DESeq(dds)
res <- results(dds, contrast = c("tumor_type", "Tubular_Adenoma", "Early_Stage_Tumor"))
```
### Clean result Table
```{r}
res_df <- as.data.frame(res)
res_df$gene_id <- rownames(res_df)

res_df <- res_df[!is.na(res_df$padj) & !is.na(res_df$log2FoldChange), ]
```
### Add gene symbols and names
```{r}
res_df$gene_symbol <- mapIds(
  org.Hs.eg.db,
  keys = res_df$gene_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

res_df$gene_name <- mapIds(
  org.Hs.eg.db,
  keys = res_df$gene_id,
  column = "GENENAME",
  keytype = "ENSEMBL",
  multiVals = "first"
)
```
### DE Summary Counts
```{r}
genes_tested <- nrow(res_df)
sig_count <- sum(res_df$padj < 0.05, na.rm = TRUE)
up_count <- sum(res_df$padj < 0.05 & res_df$log2FoldChange > 1, na.rm = TRUE)
down_count <- sum(res_df$padj < 0.05 & res_df$log2FoldChange < -1, na.rm = TRUE)

cat("Genes tested:", genes_tested, "\n")
cat("Significant genes (FDR < 0.05):", sig_count, "\n")
cat("Upregulated genes:", up_count, "\n")
cat("Downregulated genes:", down_count, "\n")
```
### Regulation groups
```{r}
res_df$Regulation <- "NS"
res_df$Regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$Regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"
```
### Top 10 genes for labeling
```{r}
top_genes <- res_df[order(res_df$padj), ][1:10, ]

top_genes$label <- ifelse(
  is.na(top_genes$gene_symbol) | top_genes$gene_symbol == "",
  top_genes$gene_id,
  top_genes$gene_symbol
)
```
### Export DEG results
```{r}
deg_results <- res_df[, c("gene_id", "log2FoldChange", "pvalue", "padj", "Regulation", "gene_symbol", "gene_name")]
colnames(deg_results) <- c(
  "Gene_ID", "log2FC", "p_value", "adjusted_p_value",
  "Regulation", "Gene_Symbol", "Gene_Name"
)

write.csv(
  deg_results,
  file.path(output_dir, "DEG_results_tubular_vs_early_stage.csv"),
  row.names = FALSE
)
```
### Count significant DEGs
```{r}
num_sig <- sum(res_df$padj < 0.05, na.rm = TRUE)
up_genes <- sum(res_df$padj < 0.05 & res_df$log2FoldChange > 0, na.rm = TRUE)
down_genes <- sum(res_df$padj < 0.05 & res_df$log2FoldChange < 0, na.rm = TRUE)
```
### Volcano plot
```{r}
volcano_plot <- ggplot(
  res_df,
  aes(x = log2FoldChange, y = -log10(padj), color = Regulation)
) +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    data = top_genes,
    aes(label = label),
    size = 3,
    max.overlaps = 100
  ) +
  scale_color_manual(values = c("Down" = "blue", "NS" = "gray", "Up" = "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  annotate(
    "text",
    x = max(res_df$log2FoldChange, na.rm = TRUE) * 0.35,
    y = max(-log10(res_df$padj), na.rm = TRUE) * 0.95,
    label = paste0(
      "Significant DEGs: ", num_sig, "\n",
      "Upregulated: ", up_genes, "\n",
      "Downregulated: ", down_genes
    ),
    hjust = 0,
    size = 4
  ) +
  theme_minimal() +
  labs(
    title = "Volcano Plot: Tubular Adenoma vs Early Stage Tumor",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value"
  )

ggsave(
  filename = file.path(output_dir, "volcano_plot_tubular_vs_early_stage.pdf"),
  plot = volcano_plot,
  width = 8,
  height = 6
)
```
### MA plot
```{r}
ma_plot <- ggplot(
  res_df,
  aes(x = baseMean, y = log2FoldChange, color = Regulation)
) +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    data = top_genes,
    aes(label = label),
    size = 3,
    max.overlaps = 100
  ) +
  scale_x_log10() +
  scale_color_manual(values = c("Down" = "blue", "NS" = "gray", "Up" = "red")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "MA Plot: Tubular Adenoma vs Early Stage Tumor",
    x = "Mean Expression (log scale)",
    y = "Log2 Fold Change"
  )

ggsave(
  filename = file.path(output_dir, "ma_plot_tubular_vs_early_stage.pdf"),
  plot = ma_plot,
  width = 8,
  height = 6
)
```
### Top 10 DEG table
```{r}
top10_deg <- head(deg_results[order(deg_results$adjusted_p_value), ], 10)
print(top10_deg)

write.csv(
  top10_deg,
  file.path(output_dir, "top10_DEGs_tubular_vs_early_stage.csv"),
  row.names = FALSE
)
```
# Early Stage Tumor vs Normal Tissue
### Set output directory for this analysis
```{r}
output_dir <- here("output", "early_stage_vs_normal")
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```
### Differential expression analysis using DESeq2
```{r}
dds <- DESeq(dds)
res <- results(dds, contrast = c("tumor_type", "Early_Stage_Tumor", "Normal"))
```
### Clean result Table
```{r}
res_df <- as.data.frame(res)
res_df$gene_id <- rownames(res_df)

res_df <- res_df[!is.na(res_df$padj) & !is.na(res_df$log2FoldChange), ]
```
### Add gene symbols and names`
```{r}
res_df$gene_symbol <- mapIds(
  org.Hs.eg.db,
  keys = res_df$gene_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

res_df$gene_name <- mapIds(
  org.Hs.eg.db,
  keys = res_df$gene_id,
  column = "GENENAME",
  keytype = "ENSEMBL",
  multiVals = "first"
)
```
### DE Summary Counts
```{r}
early_genes_tested <- nrow(res_df)
early_sig_count <- sum(res_df$padj < 0.05, na.rm = TRUE)
early_up_count <- sum(res_df$padj < 0.05 & res_df$log2FoldChange > 1, na.rm = TRUE)
early_down_count <- sum(res_df$padj < 0.05 & res_df$log2FoldChange < -1, na.rm = TRUE)

cat("Genes tested:", early_genes_tested, "\n")
cat("Significant genes (FDR < 0.05):", early_sig_count, "\n")
cat("Upregulated genes:", early_up_count, "\n")
cat("Downregulated genes:", early_down_count, "\n")
```
### Regulation groups
```{r}
res_df$Regulation <- "NS"
res_df$Regulation[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$Regulation[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"
```
### Top 10 genes for labeling
```{r}
top_genes <- res_df[order(res_df$padj), ][1:10, ]

top_genes$label <- ifelse(
  is.na(top_genes$gene_symbol) | top_genes$gene_symbol == "",
  top_genes$gene_id,
  top_genes$gene_symbol
)
```
### Export DEG results
```{r}
deg_results <- res_df[, c("gene_id", "log2FoldChange", "pvalue", "padj", "Regulation", "gene_symbol", "gene_name")]
colnames(deg_results) <- c(
  "Gene_ID", "log2FC", "p_value", "adjusted_p_value",
  "Regulation", "Gene_Symbol", "Gene_Name"
)
write.csv(
  deg_results,
  file.path(output_dir, "DEG_results_early_stage_vs_normal.csv"),
  row.names = FALSE
)
```
### Count significant DEGs
```{r}
num_sig <- sum(res_df$padj < 0.05, na.rm = TRUE)
up_genes <- sum(res_df$padj < 0.05 & res_df$log2FoldChange > 0, na.rm = TRUE)
down_genes <- sum(res_df$padj < 0.05 & res_df$log2FoldChange < 0, na.rm = TRUE)
```
### Volcano plot
```{r}
volcano_plot_early_stage <- ggplot(
  res_df,
  aes(x = log2FoldChange, y = -log10(padj), color = Regulation)
) +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    data = top_genes,
    aes(label = label),
    size = 3,
    max.overlaps = 100
  ) +
  scale_color_manual(values = c("Down" = "blue", "NS" = "gray", "Up" = "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  annotate(
  "text",
  x = 1.5,
  y = 26,
  label = paste0(
    "Significant DEGs: ", early_sig_count, "\n",
    "Upregulated: ", early_up_count, "\n",
    "Downregulated: ", early_down_count
  ),
  hjust = 0,
  vjust = 1,
  size = 3.6
) +
coord_cartesian(clip = "off") +
theme(
  plot.margin = margin(10, 40, 10, 10)
) +
  theme_minimal() +
  labs(
    title = "Volcano Plot: Early Stage Tumor vs Normal",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value"
  )

ggsave(
  filename = file.path(output_dir, "volcano_plot_early_stage_vs_normal.pdf"),
  plot = volcano_plot_early_stage,
  width = 8,
  height = 6
)
```
### MA plot
```{r}
ma_plot_early_stage <- ggplot(
  res_df,
  aes(x = baseMean, y = log2FoldChange, color = Regulation)
) +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    data = top_genes,
    aes(label = label),
    size = 3,
    max.overlaps = 100
  ) +
  scale_x_log10() +
  scale_color_manual(values = c("Down" = "blue", "NS" = "gray", "Up" = "red")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "MA Plot: Early Stage Tumor vs Normal",
    x = "Mean Expression (log scale)",
    y = "Log2 Fold Change"
  )

ggsave(
  filename = file.path(output_dir, "ma_plot_early_stage_vs_normal.pdf"),
  plot = ma_plot_early_stage,
  width = 8,
  height = 6
)
```
### Top 10 DEG table
```{r}
top10_deg <- head(deg_results[order(deg_results$adjusted_p_value), ], 10)
print(top10_deg)

write.csv(
  top10_deg,
  file.path(output_dir, "top10_DEGs_early_stage_vs_normal.csv"),
  row.names = FALSE
)
```
# Upset Plot
### Create output directory for upset plot
```{r}
output_dir <- here("output", "upset_plot_all_groups")
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```
### Read DEG result files
```{r}
villous <- read.csv(
  here("output", "villous_vs_normal", "DEG_results_villous_vs_normal.csv"),
  stringsAsFactors = FALSE
)

tubular <- read.csv(
  here("output", "tubular_vs_normal", "DEG_results_tubular_vs_normal.csv"),
  stringsAsFactors = FALSE
)

early_stage <- read.csv(
  here("output", "early_stage_vs_normal", "DEG_results_early_stage_vs_normal.csv"),
  stringsAsFactors = FALSE
)

advanced <- read.csv(
  here("output", "advanced_vs_normal", "DEG_results_advanced_vs_normal.csv"),
  stringsAsFactors = FALSE
)
```
### Filter significant DEGs, use same rule for all groups
```{r}
villous_genes <- unique(
  villous$Gene_ID[
    !is.na(villous$adjusted_p_value) &
    !is.na(villous$log2FC) &
    villous$adjusted_p_value < 0.05 &
    abs(villous$log2FC) > 1
  ]
)

tubular_genes <- unique(
  tubular$Gene_ID[
    !is.na(tubular$adjusted_p_value) &
    !is.na(tubular$log2FC) &
    tubular$adjusted_p_value < 0.05 &
    abs(tubular$log2FC) > 1
  ]
)

early_stage_genes <- unique(
  early_stage$Gene_ID[
    !is.na(early_stage$adjusted_p_value) &
    !is.na(early_stage$log2FC) &
    early_stage$adjusted_p_value < 0.05 &
    abs(early_stage$log2FC) > 1
  ]
)

advanced_genes <- unique(
  advanced$Gene_ID[
    !is.na(advanced$adjusted_p_value) &
    !is.na(advanced$log2FC) &
    advanced$adjusted_p_value < 0.05 &
    abs(advanced$log2FC) > 1
  ]
)
```
### Print DEG counts
```{r}
cat("Villous vs Normal DEGs:", length(villous_genes), "\n")
cat("Tubular vs Normal DEGs:", length(tubular_genes), "\n")
cat("Early Stage vs Normal DEGs:", length(early_stage_genes), "\n")
cat("Advanced vs Normal DEGs:", length(advanced_genes), "\n")
```
### Build list for UpSet plot
```{r}
deg_list <- list(
  Villous = villous_genes,
  Tubular = tubular_genes,
  EarlyStage = early_stage_genes,
  Advanced = advanced_genes
)
```
### Convert to UpSet input format
```{r}
upset_data <- fromList(deg_list)
```
### Save the binary membership table too 
```{r}
write.csv(
  upset_data,
  file.path(output_dir, "upset_membership_table.csv"),
  row.names = TRUE
)
```
### Plot UpSet
```{r}
pdf(file.path(output_dir, "upset_plot_all_groups.pdf"), width = 10, height = 6)

print(
  upset(
    upset_data,
    nsets = 5,
    nintersects = 20,
    order.by = "freq"
  )
)

dev.off()

upset(
  upset_data,
  nsets = 4,
  nintersects = NA,
  order.by = "freq",
  mainbar.y.label = "Shared DEG Intersections",
  sets.x.label = "DEGs per Comparison",
  text.scale = 1.3
)

dev.off()

cat("UpSet plot saved to:", file.path(output_dir, "upset_plot_all_groups.pdf"), "\n")
cat("Membership table saved to:", file.path(output_dir, "upset_membership_table.csv"), "\n")
```
### Create subfolder for multi-panel figures
```{r}
if (!dir.exists("output/multi_panel")) {
  dir.create("output/multi_panel")
}
```
### Volcano Multi-Panel Figure 
```{r}
volcano_plot_advanced <- volcano_plot_advanced + ggtitle("Advanced vs Normal")
volcano_plot_early_stage <- volcano_plot_early_stage + ggtitle("Early-Stage vs Normal")
volcano_plot_tubular <- volcano_plot_tubular + ggtitle("Tubular Adenoma vs Normal")
volcano_plot_villous <- volcano_plot_villous + ggtitle("Villous Adenoma vs Normal")
```
### Combine volcano plots into 2x2 panel 
```{r}
volcano_panel <- (volcano_plot_advanced | volcano_plot_early_stage) /
                 (volcano_plot_tubular | volcano_plot_villous)
```
### Add panel labels (A, B, C, D)
```{r}
volcano_panel <- volcano_panel +
  plot_annotation(tag_levels = "A")
```
### Display
```{r}
volcano_panel
```
### Save multi-panel volcano figure
```{r}
ggsave("output/multi_panel/volcano_panel.pdf",
       volcano_panel,
       width = 10,
       height = 8,
       dpi = 300)
```
### MA Multi-Panel Figure
```{r}
ma_plot_advanced <- ma_plot_advanced + ggtitle("Advanced vs Normal")
ma_plot_early_stage <- ma_plot_early_stage + ggtitle("Early-Stage vs Normal")
ma_plot_tubular <- ma_plot_tubular + ggtitle("Tubular Adenoma vs Normal")
ma_plot_villous <- ma_plot_villous + ggtitle("Villous Adenoma vs Normal")
```
### Combine MA plots into 2x2 panel
```{r}
ma_panel <- (ma_plot_advanced | ma_plot_early_stage) /
            (ma_plot_tubular | ma_plot_villous)
```
### Add panel labels (A, B, C, D)
```{r}
ma_panel <- ma_panel +
  plot_annotation(tag_levels = "A")
```
### Display
```{r}
ma_panel
```
### Save multi-panel MA figure
```{r}
ggsave("output/multi_panel/ma_panel.pdf",
       ma_panel,
       width = 10,
       height = 8,
       dpi = 300)
```
# Count Bar Plot
### Create DEG counts dataframe
```{r}
deg_counts <- data.frame(
  Tumor = c("Advanced Tumor", "Early-Stage Tumor", "Tubular Adenoma", "Villous Adenoma"),
  Up = c(advanced_up_count, early_up_count, tubular_up_count, villous_up_count),
  Down = c(advanced_down_count, early_down_count, tubular_down_count, villous_down_count)
)
```
### Convert to long format
```{r}
deg_counts_long <- deg_counts %>%
  pivot_longer(cols = c("Up", "Down"),
               names_to = "Regulation",
               values_to = "Count")
```
### Total counts for labels
```{r}
deg_totals <- deg_counts_long %>%
  group_by(Tumor) %>%
  summarise(Total = sum(Count), .groups = "drop")
```
### Plot 
```{r}
deg_count_barplot <- ggplot(deg_counts_long, aes(x = Tumor, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Up" = "red", "Down" = "blue")) +
   geom_text(data = deg_totals,
            aes(x = Tumor, y = Total, label = Total),
            inherit.aes = FALSE,
            vjust = -0.4,
            size = 4) +
  
  theme_minimal() +
  labs(
    title = "Differentially Expressed Genes Across Tumor Types",
    x = "Tumor Type",
    y = "Number of DEGs"
  ) +
  theme(
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```
### Display and save
```{r}
deg_count_barplot

ggsave(
  filename = "output/count_bar_plot/deg_count_barplot.png",
  plot = deg_count_barplot,
  width = 8,
  height = 6,
  dpi = 300
)

ggsave(
  filename = "output/count_bar_plot/deg_count_barplot.pdf",
  plot = deg_count_barplot,
  width = 8,
  height = 6
)
```